---
title: "Causal Final Project - Propensity Score Matching"
output:
  pdf_document: default
  html_document: default
\usepackage{float}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```
   
```{r, echo=FALSE, message=FALSE}
require(dplyr)
require(MASS)
require(DescTools)
require(optmatch) 
require(Matching)
require(arm)
require(knitr)
require(Hmisc)
```
  
   
Note: data is from: https://www.kaggle.com/janiobachmann/math-students/data   

For this project, I am imagining that students are recruited for a math tutoring program, which provides supplemental support throughout a single school year. Demographic information is available for each student, including a pre-test score, and I am interested in measuring the effect of this program on students' final grades in their math course.  

In order to make this data look more realistic, I have pulled a dataset from Kaggle, which includes information on a group of high school math students. Because there are so many variables, I first used a stepAIC algorithm to get a sense of which variables were significant predictors (in real life) of students' final test scores. Ten variables were selected by this algorithm, and I chose to keep these plus two others (sex and health status) as my variable pool for this simulation.   
```{r, cache=TRUE}
data <- read.csv("./Data/student-mat.csv")
data <- data %>% dplyr::filter(school=="GP")
data <- data %>% dplyr::select(sex, age, address, Pstatus, Medu, 
                               Fedu, traveltime, studytime, romantic,
                               failures, schoolsup, famsup, schoolsup, 
                               activities, nursery, higher,internet, 
                               famrel, freetime, goout, Dalc, Walc, health, 
                               absences, G1, G3)
null_model <- lm(G3 ~ 1, data = data)
full_model <- lm(G3 ~ ., data = data)
AIC_model <- stepAIC(null_model, 
                     scope = list(upper = full_model, lower = null_model), 
                     direction = "forward", trace = FALSE)
summary(AIC_model)
data <- data %>% dplyr::select(G3, G1, sex, age, romantic, absences, failures, 
                               schoolsup, Walc, traveltime, famrel, goout, health)

# this is to help me build the logit model for my hypothetical "treatment"
get_sch_supp <- glm(schoolsup ~ . , data = data[,-1], family = binomial(link = "logit"))
summary(get_sch_supp) 

data <- data[,-which(names(data)=="schoolsup")]
```   
    
Data Dictionary:    
sex - student's sex (binary: 'F' - female or 'M' - male)  
age - student's age (numeric: from 15 to 22)  
traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)  
failures - number of past class failures (numeric: n if 1<=n<3, else 4)  
schoolsup - extra educational support (binary: yes or no)  
romantic - in a romantic relationship (binary: yes or no)
famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)  
goout - going out with friends (numeric: from 1 - very low to 5 - very high)  
Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)  
absences - number of school absences (numeric: from 0 to 93)  
health - current health status (numeric: from 1 - very bad to 5 - very good)  
G1 - first period grade in Math class (scored from 0-20)  
  
Now, I am going to generate data under the following two models:    

## Data generation 1 (World A)
World A: linear relationship where all variables predict both treatment assignment and potential outcomes.  
```{r, cache=TRUE} 
set.seed(123)

dataGen_A <- function(data){
  treat = 9 - 0.41 * data$G1 - 0.4 * data$age - 
    0.25 * (data$romantic=="yes") - 0.03 * data$absences + 
    0.3 * data$failures - 0.1 * data$Walc - 0.1 * data$traveltime + 
    0.2 * data$famrel - 0.05 * data$goout + 0.4 * data$health - 
    0.5 * (data$sex == "M") +
    rnorm(nrow(data), 0, 1)

  prob = LogitInv(treat)
  treat = rbinom(nrow(data), 1, prob=prob)

  Y0 = 2.9 + 0.9 * data$G1 + 0.2 * data$age - 
    0.5 * (data$romantic=="yes") + 
    0.15 * data$absences - 0.9 * data$failures - 
    0.8 * data$Walc - 1 * data$traveltime + 
    0.5 * data$famrel - 0.5 * data$goout + 
    0.2 * data$health - 0.5 * (data$sex == "M") +
    rnorm(nrow(data), 0, 1)

  Y1 = 2.9 + 0.9 * data$G1 + 0.2 * data$age - 
    0.5 * (data$romantic=="yes") + 
    0.15 * data$absences - 0.9 * data$failures - 
    0.8 * data$Walc - 1 * data$traveltime + 
    0.5 * data$famrel - 0.5 * data$goout + 
    0.2 * data$health - 0.5 * (data$sex == "M") + 
    3 + rnorm(nrow(data), 0, 1)

  Y = ifelse(treat == 1, Y1, Y0)
  dataA = data.frame(Y=Y,treat=treat)
  covars = data[,-1]
  dataA = cbind(dataA, covars)
  return(list(dataA, data.frame(Y0=Y0,Y1=Y1)))
}

#generate some data
dat <- dataGen_A(data)
dataA <- dat[[1]]
outcomes <- dat[[2]]

#calculate SATT
SATT_A <- mean(outcomes$Y1[dataA$treat==1]) - mean(outcomes$Y0[dataA$treat==1])
SATT_A
```
  
## Data generation 2 (World B) 
World B: Make the response surface more complex by adding interation and quadratic terms and making treatment effect moderated by pre-test score.    
```{r, cache=TRUE}  
set.seed(111)


dataGen_B <- function(data){
  treat = 4 - 0.3 * data$G1 + .021 * data$G1^2 - 0.3 * data$age - 
    0.1 * data$age * (data$sex=="M") - 0.25 * (data$romantic=="yes") - 
    0.03 * data$absences + 0.3 * data$failures - 0.1 * data$failures^2 - 
    0.1 * data$Walc - 0.12 * data$traveltime + 0.2 * data$famrel - 
    0.05 * data$goout + 0.1 * data$health + 0.05 * data$health^2 - 
    0.3 * (data$sex == "M") -.12 * data$famrel * data$traveltime + 
    .08 * data$famrel * data$health + rnorm(nrow(data), 0, 1)

  prob = LogitInv(treat)
  treat = rbinom(nrow(data), 1, prob=prob)

  Y0 = 14 + 1.3 * data$G1 - .07 * data$G1^2 + 0.1 * data$age - 
    0.5 * data$age * (data$sex=="M") - 0.5 * (data$romantic=="yes") + 
    0.1 * data$absences - 0.5 * data$failures - 0.3 * data$Walc - 
    1 * data$traveltime + 0.5 * data$famrel - 0.2 * data$goout + 
    0.2 * data$health - 0.5 * (data$sex == "M") - 0.1 * data$failures^2 - 
    .2 * data$famrel * data$traveltime - .1 * data$famrel * data$health + 
    + 0.05 * data$health^2 + rnorm(nrow(data), 0, 1)

  Y1 = 14 + 1.3 * data$G1 - .07 * data$G1^2 + 0.1 * data$age - 
    0.5 * data$age * (data$sex=="M") - 0.5 * (data$romantic=="yes") + 
    0.1 * data$absences - 0.5 * data$failures - 0.3 * data$Walc - 
    1 * data$traveltime + 0.5 * data$famrel - 0.2 * data$goout + 
    0.2 * data$health - 0.5 * (data$sex == "M") - 0.1 * data$failures^2 - 
    .2 * data$famrel * data$traveltime - .1 * data$famrel * data$health + 
    0.05 * data$health^2 + rnorm(nrow(data), 0, 1) + 4 - .1 * data$G1

  Y = ifelse(treat == 1, Y1, Y0)
  dataB = data.frame(Y=Y,treat=treat)
  covars = data[,-1]
  dataB = cbind(dataB, covars)
  
  return(list(dataB, data.frame(Y0=Y0,Y1=Y1)))
}

#generate some data
dat <- dataGen_B(data)
dataB <- dat[[1]]
outcomes <- dat[[2]]

#look at overlap of propensity scores for this data
pscore_fit3 <- glm(treat ~ . + I(G1^2) + age:sex + I(failures^2) 
                   + I(health^2) + famrel:traveltime , 
                   data = dataB[,-c(1)], family = "binomial")
pscore3 <- predict(pscore_fit3, dataB[-1], type = "response")
hist(pscore3[dataB$treat==0], col = rgb(0.9,0.8,0.0,0.5), xlim=c(0,1))
hist(pscore3[dataB$treat==1], add=TRUE, col = rgb(0.0,0.5,0.9,0.5))

#calculate SATT
SATT_B <- mean(outcomes$Y1[dataB$treat==1]) - mean(outcomes$Y0[dataB$treat==1])
SATT_B
```
     
Functions:  
```{r, cache=TRUE}
get_IPTW_weights <- function(treat, pscores){
  weights <- rep(0, length(treat))
  weights[treat==1] <- 1
  weights[treat==0] <- pscores[treat==0]/(1-pscores[treat==0])
  nc = sum(treat==0)
  swc = sum(weights[treat==0])
  weights[treat==0] <- (weights[treat==0]/swc)*nc
  return(weights)
}

get_optm_weights <- function(treat, matches){
  chosen_controls <- matches[treat==0] %>% na.omit() %>% names() %>% as.numeric()
  weights <- rep(0, length(treat))
  weights[chosen_controls] <- 1
  weights[treat==1] <- 1
  return(weights)
}

get_gen_weights <- function(treat, matches){
  weights <- rep(0, length(treat))
  matched_controls <- unique(matches[,2])
  control_weights <- rep(0, length(matched_controls))
  for(i in 1:length(matched_controls)){
    weight_list <- matches[,3][matches[,2]==matched_controls[i]]
    control_weights[i] <- sum(weight_list)
  }
  weights[matched_controls] <- control_weights
  weights[treat==1] <- 1
  return(weights)
}

balance_func = function(df, covariates, weights){
  ncov = length(covariates)
  mn1 = mn0 = mn1.m = mn0.m = diff = diff.m = ratio = ratio.m = rep(0, ncov)
  res = data.frame(mn1, mn0, mn1.m, mn0.m, diff, diff.m, ratio, ratio.m)
  rownames(res) = covariates
  w0 <- weights[which(df$treat==0)]
  w1 <- weights[which(df$treat==1)]
  for(i in 1:ncov){
    res[i,"mn1"] = mean(df[,(covariates[i])][df$treat==1], na.rm=T)
    res[i,"mn0"] = mean(df[,(covariates[i])][df$treat==0], na.rm=T)
    res[i,"mn1.m"] = sum(df[,(covariates[i])][df$treat==1]*w1)/sum(w1)
    res[i,"mn0.m"] = sum(df[,(covariates[i])][df$treat==0]*w0)/sum(w0)
    if(length(unique(df[,(covariates[i])])) > 2){
      res[i,"diff"] = (res[i,"mn1"] - 
                         res[i,"mn0"])/sd(df[,(covariates[i])][df$treat==1])
      res[i,"diff.m"] = (res[i,"mn1.m"] - 
                           res[i,"mn0.m"])/sd(df[,(covariates[i])][df$treat==1])
      res[i,"ratio"] = sd(df[,(covariates[i])][df$treat==0])/
        sd(df[,(covariates[i])][df$treat==1])
      res[i,"ratio.m"] = sqrt(Hmisc::wtd.var(df[,(covariates[i])][df$treat==0], w0))/
        sqrt(Hmisc::wtd.var(df[,(covariates[i])][df$treat==1], w1))
    } else {
      res[i,"diff"] = res[i,"mn1"] - res[i,"mn0"]
      res[i,"diff.m"] = res[i,"mn1.m"] - res[i,"mn0.m"]
      res[i,"ratio"] = 0
      res[i,"ratio.m"] = 0 
      }
  }
  return(round(res,3))
}
```


#  Model fitting for World A: The correct model  

Calculate p-scores  
```{r, cache=TRUE}
pscore_fit1 <- glm(treat ~ . , data = dataA[,-c(1)], family = "binomial")
pscore1 <- predict(pscore_fit1, dataA[-1], type = "response")  
```
   
IPTW weights   
```{r, cache=TRUE}
weightsm1 <- get_IPTW_weights(dataA$treat, pscore1)
est_1 <- lm(Y ~ . , weights=weightsm1, data = dataA)
est_1 <- summary(est_1)$coef["treat",1]
est_1

XcovA <- dataA
XcovA$sex <- as.numeric(XcovA$sex=="M")
XcovA$romantic <- as.numeric(XcovA$romantic=="yes")

covsA <- c("G1", "sex", "age", "romantic", "absences", "failures", "Walc", "traveltime", "famrel", "goout", "health")

balance_func(XcovA, covsA, weightsm1)
```
    
  

Optimal matching  
```{r, cache=TRUE}
m2 <- pairmatch(pscore_fit1)
weightsm2 <- get_optm_weights(dataA$treat, m2)
est_2 <- lm(Y ~ . , weights=weightsm2, data = dataA)
est_2 <- summary(est_2)$coef["treat",1]
est_2

balance_func(XcovA, covsA, weightsm2)
```
  
Genetic matching  
```{r, warning=FALSE, message=FALSE, cache=TRUE}
#save just the x covariates
Xcov <- dataA[,-c(1:2)]

#add the propensity scores  
Xcov$pscore <- pscore1

#change factor variables to 1s and 0s
Xcov$romantic <- as.numeric(Xcov$romantic=="yes")
Xcov$sex <- as.numeric(Xcov$sex=="M")

#run genetic matching
m3 <- GenMatch(dataA$treat, Xcov, BalanceMatrix=Xcov, 
               estimand="ATT", pop.size = 100, print=FALSE)

#create weights based on genetic matching
weightsm3 <- get_gen_weights(dataA$treat, m3$matches)

#estimate treatment effect using the weights
est_3 <- lm(Y ~ . , weights=weightsm3, data = dataA)
est_3 <- summary(est_3)$coef["treat",1]
est_3

# Not sure why this is giving me a different answer:
mgen <- Match(Y = dataA$Y, Tr = dataA$treat, X = Xcov, Weight.matrix = m3)
mgen$est

# Actually: this is the same as a regression without the other covariates
summary(lm(Y ~ treat , weights=weightsm3, data = dataA))$coef["treat",1]

balance_func(XcovA, covsA, weightsm3)
```
  
#  Model fitting for World A: Missing some covariates   

Calculate p-scores  
```{r, cache=TRUE}
pscore_fit2 <- glm(treat ~ G1 + sex + age + failures + absences, 
                   data = dataA, family = "binomial")
pscore2 <- predict(pscore_fit2, dataA, type = "response")
```
   
IPTW weights   
```{r, cache=TRUE}
weightsm4 <- get_IPTW_weights(dataA$treat, pscore2)
est_4 <- lm(Y ~ treat + G1 + sex + age + failures + absences, weights=weightsm4, data = dataA)
est_4 <- summary(est_4)$coef["treat",1]
est_4
```
    
Optimal matching  
```{r, cache=TRUE}
m5 <- pairmatch(pscore_fit2)
weightsm5 <- get_optm_weights(dataA$treat, m5)
est_5 <- lm(Y ~ treat + G1 + sex + age + failures + absences, weights=weightsm5, data = dataA)
est_5 <- summary(est_5)$coef["treat",1]
est_5
```
  
Genetic matching  
```{r, warning=FALSE, message=FALSE, cache=TRUE}
#save shorter covariate list
Xcov <- Xcov %>% dplyr::select(G1, sex, age, failures, absences)
Xcov$pscore <- pscore2

#run genetic matching
m6 <- GenMatch(dataA$treat, Xcov, BalanceMatrix=Xcov, 
               estimand="ATT", pop.size = 100, print=FALSE)

#create weights based on genetic matching
weightsm6 <- get_gen_weights(dataA$treat, m6$matches)

#estimate treatment effect using the weights
est_6 <- lm(Y ~ treat + G1 + sex + age + failures + absences, 
            weights=weightsm6, data = dataA)
est_6 <- summary(est_6)$coef["treat",1]
est_6
```  
    
Verify that the treatment group is never larger than controls  
```{r}
ntreat <- rep(NA, 1000)
for(i in 1:1000){
  dataA <- dataGen_A(data)[[1]]
  ntreat[i] <- sum(dataA$treat==1)
}
sum((ntreat/nrow(dataA))>.5)
mean(ntreat/nrow(dataA))
```  
  
Randomization distributions (note: this takes a very long time to run):  
```{r, cache=TRUE, message=FALSE, warning=FALSE}
nSim  <- 1000
simA_IPTW <- simA_IPTW_WrongMod <- rep(NA, nSim)
simA_optm <- simA_optm_WrongMod <- rep(NA, nSim)
simA_Gen <- simA_Gen_WrongMod <- rep(NA, nSim)

mat <- matrix(NA, ncol=length(covsA), nrow=nSim)
colnames(mat) <- covsA
simA_IPTW_bal <- simA_IPTW_WrongMod_bal <- mat
simA_optm_bal <- simA_optm_WrongMod_bal <- mat
simA_Gen_bal <- simA_Gen_WrongMod_bal <- mat

for(i in 1:1000){
  dataA <- dataGen_A(data)[[1]]
  
  #pscores
  pscore_fit1 <- glm(treat ~ . , data = dataA[,-c(1)], family = "binomial")
  pscore1 <- predict(pscore_fit1, dataA[-1], type = "response")
  pscore_fit2 <- glm(treat ~ G1 + sex + age + failures + absences, 
                   data = dataA, family = "binomial")
  pscore2 <- predict(pscore_fit2, dataA, type = "response")
  
  #iptw
  weightsm1 <- get_IPTW_weights(dataA$treat, pscore1)
  est_1 <- lm(Y ~ . , weights=weightsm1, data = dataA)
  simA_IPTW[i] <- summary(est_1)$coef["treat",1]
  simA_IPTW_bal[i,] <- balance_func(XcovA, covsA, weightsm1)$diff.m
  
  #optimal
  m2 <- pairmatch(pscore_fit1)
  weightsm2 <- get_optm_weights(dataA$treat, m2)
  est_2 <- lm(Y ~ . , weights=weightsm2, data = dataA)
  simA_optm[i] <- summary(est_2)$coef["treat",1]
  simA_optm_bal[i,] <- balance_func(XcovA, covsA, weightsm2)$diff.m
  
  #save just the x covariates
  Xcov <- XcovA[,-c(1:2)]
  #add the propensity scores  
  #Xcov$pscore <- pscore1

  #genetic
  m3 <- GenMatch(dataA$treat, Xcov, BalanceMatrix=Xcov, 
               estimand="ATT", pop.size = 100, print=FALSE)
  weightsm3 <- get_gen_weights(dataA$treat, m3$matches)
  est_3 <- lm(Y ~ . , weights=weightsm3, data = dataA)
  simA_Gen[i] <- summary(est_3)$coef["treat",1]
  simA_Gen_bal[i,] <- balance_func(XcovA, covsA, weightsm3)$diff.m
  
  #iptw
  weightsm4 <- get_IPTW_weights(dataA$treat, pscore2)
  est_4 <- lm(Y ~ treat + G1 + sex + age + failures + absences, 
              weights=weightsm4, data = dataA)
  simA_IPTW_WrongMod[i] <- summary(est_4)$coef["treat",1]
  simA_IPTW_WrongMod_bal[i,] <- balance_func(XcovA, covsA, weightsm4)$diff.m
  
  #optimal
  m5 <- pairmatch(pscore_fit2)
  weightsm5 <- get_optm_weights(dataA$treat, m5)
  est_5 <- lm(Y ~ treat + G1 + sex + age + failures + absences, 
              weights=weightsm5, data = dataA)
  simA_optm_WrongMod[i] <- summary(est_5)$coef["treat",1]
  simA_optm_WrongMod_bal[i,] <- balance_func(XcovA, covsA, weightsm5)$diff.m
  
  #save less covars
  Xcov <- Xcov %>% dplyr::select(G1, sex, age, failures, absences)
  #Xcov$pscore <- pscore2

  #run genetic matching
  m6 <- GenMatch(dataA$treat, Xcov, BalanceMatrix=Xcov, 
               estimand="ATT", pop.size = 100, print=FALSE)
  weightsm6 <- get_gen_weights(dataA$treat, m6$matches)

  #estimate treatment effect using the weights
  est_6 <- lm(Y ~ treat + G1 + sex + age + failures + absences, 
            weights=weightsm6, data = dataA)
  simA_Gen_WrongMod[i] <- summary(est_6)$coef["treat",1]
  simA_Gen_WrongMod_bal[i,] <- balance_func(XcovA, covsA, weightsm6)$diff.m
  
  print(i)
  
}
```    
   
Save everything, just in case!
```{r}
# statsA_final <-cbind(simA_IPTW,  simA_optm,  simA_Gen, simA_IPTW_WrongMod,
# simA_optm_WrongMod,simA_Gen_WrongMod)
# 
# colnames(statsA_final) <- c("simA_IPTW","simA_optm","simA_Gen","simA_IPTW_WrongMod",
# "simA_optm_WrongMod","simA_Gen_WrongMod")
# 
# write.csv(statsA_final, "./statsA_final.csv")
# write.csv(simA_IPTW_bal, "./IPTW_bal.csv")
# write.csv(simA_optm_bal, "./optm_bal.csv")
# write.csv(simA_Gen_bal, "./Gen_bal.csv")
# write.csv(simA_IPTW_WrongMod_bal, "./IPTW_WM_bal.csv")
# write.csv(simA_optm_WrongMod_bal, "./optm_WM_bal.csv")
# write.csv(simA_Gen_WrongMod_bal, "./Gen_WM_bal.csv")
```
    
      
Summary table:  
```{r, cache=TRUE}
Biases <- round(abs(c(mean(simA_IPTW),
                     mean(simA_optm),
                     mean(simA_Gen),
                     mean(simA_IPTW_WrongMod),
                     mean(simA_optm_WrongMod),
                     mean(simA_Gen_WrongMod))-3),2)
RMSEs <- round(c(sqrt(mean((simA_IPTW-3)^2)),
                 sqrt(mean((simA_optm-3)^2)),
                 sqrt(mean((simA_Gen-3)^2)),
                 sqrt(mean((simA_IPTW_WrongMod-3)^2)),
                 sqrt(mean((simA_optm_WrongMod-3)^2)),
                 sqrt(mean((simA_Gen_WrongMod-3)^2))),2)
Included_Diff <- round(c(mean(abs(simA_IPTW_bal[,c(1:11)])),
                     mean(abs(simA_optm_bal[,c(1:11)])),
                     mean(abs(simA_Gen_bal[,c(1:11)])),
                     mean(abs(simA_IPTW_WrongMod_bal[,c(1:3,5:6)])),
                     mean(abs(simA_optm_WrongMod_bal[,c(1:3,5:6)])),
                     mean(abs(simA_Gen_WrongMod_bal[,c(1:3,5:6)]))),2)
Excluded_Diff <- round(c(mean(abs(simA_IPTW_bal[,c(1:11)])),
                     mean(abs(simA_optm_bal[,c(1:11)])),
                     mean(abs(simA_Gen_bal[,c(1:11)])),
                     mean(abs(simA_IPTW_WrongMod_bal[,c(4,7:11)])),
                     mean(abs(simA_optm_WrongMod_bal[,c(4,7:11)])),
                     mean(abs(simA_Gen_WrongMod_bal[,c(4,7:11)]))),2)

results <- data.frame(Bias = Biases,
                      RMSE = RMSEs,
                      Mean_Sd_Diff_Included = Included_Diff,
                      Mean_Sd_Diff_NotIncluded = Excluded_Diff)
rownames(results) <- c("IPTW - Correct Model", "Optimal - Correct Model", 
                       "Genetic - Correct Model",
                       "IPTW - Wrong Model", "Optimal - Wrong Model",
                       "Genetic - Wrong Model")
kable(results)
```    
  
Sampling distributions:  
```{r, cache=TRUE}
par(mfrow = c(2,3))
hist(simA_IPTW, xlim=c(2,4.5), main="IPTW Correct Model", xlab="ATT Estimate")
abline(v=3, lwd=2, col=4)
abline(v=mean(simA_IPTW), lwd=2, col=2, lty=2)
hist(simA_optm, xlim=c(2,4.5), main="Opt Matching Correct Model", xlab="ATT Estimate")
abline(v=3, lwd=2, col=4)
abline(v=mean(simA_optm), lwd=2, col=2, lty=2)
hist(simA_Gen, xlim=c(2,4.5), main="Gen Matching Correct Model", xlab="ATT Estimate")
abline(v=3, lwd=2, col=4)
abline(v=mean(simA_Gen), lwd=2, col=2, lty=2)
hist(simA_IPTW_WrongMod, xlim=c(2,4.5), main="IPTW Wrong Model", xlab="ATT Estimate")
abline(v=3, lwd=2, col=4)
abline(v=mean(simA_IPTW_WrongMod), lwd=2, col=2, lty=2)
hist(simA_optm_WrongMod, xlim=c(2,4.5), main="Opt Matching Wrong Model", xlab="ATT Estimate")
abline(v=3, lwd=2, col=4)
abline(v=mean(simA_optm_WrongMod), lwd=2, col=2, lty=2)
hist(simA_Gen_WrongMod, xlim=c(2,4.5), main="Gen Matching Wrong Model", xlab="ATT Estimate")
abline(v=3, lwd=2, col=4)
abline(v=mean(simA_Gen_WrongMod), lwd=2, col=2, lty=2)
```
     

#  Model fitting for World B: The correct model  

Calculate p-scores  
```{r, cache=TRUE}
pscore_fit3 <- glm(treat ~ . + I(G1^2) + age:sex + I(failures^2) 
                   + I(health^2) + famrel:traveltime + famrel:health, 
                   data = dataB[,-c(1)], family = "binomial")
pscore3 <- predict(pscore_fit3, dataB[-1], type = "response")

#saving x covariates
XcovB <- dataB[,-c(1)]
XcovB$romantic <- as.numeric(XcovB$romantic=="yes")
XcovB$sex <- as.numeric(XcovB$sex=="M")
XcovB$G1sq <- (XcovB$G1)^2
XcovB$agesex <- XcovB$age * XcovB$sex
XcovB$failuressq <- (XcovB$failures)^2
XcovB$healthsq <- (XcovB$health)^2
XcovB$famreltravtime <- XcovB$famrel * XcovB$traveltime
XcovB$famrelhealth <- XcovB$famrel * XcovB$health

#saving covariate names
covsB <- c("G1", "sex", "age", "romantic", "absences", "failures", "Walc", "traveltime", "famrel", "goout", "health", "G1sq", "agesex", "failuressq", "healthsq", "famreltravtime","famrelhealth")
```
   
IPTW weights   
```{r, cache=TRUE}
weightsm1_B <- get_IPTW_weights(dataB$treat, pscore3)
est_1_B <- lm(Y ~ . + I(G1^2) + age:sex + famrel:traveltime + I(failures^2) 
              + I(health^2) + famrel:health, weights=weightsm1_B, data = dataB)
est_1_B <- summary(est_1_B)$coef["treat",1]
est_1_B

balance_func(XcovB, covsB, weightsm1_B)
```
    
Optimal matching  
```{r, cache=TRUE}
m2_B <- pairmatch(treat ~ . + I(G1^2) + age:sex + I(failures^2) 
                   + I(health^2) + famrel:traveltime , 
                   data = dataB[,-c(1)])
weightsm2_B <- get_optm_weights(dataB$treat, m2_B)
est_2_B <- lm(Y ~ . + I(G1^2) + age:sex + famrel:traveltime + 
              famrel:health, weights=weightsm2_B, data = dataB)
est_2_B <- summary(est_2_B)$coef["treat",1]
est_2_B

balance_func(XcovB, covsB, weightsm2_B)
```
  
Genetic matching   
```{r, warning=FALSE, message=FALSE, cache=TRUE}
#save just the x covariates
Xcov <- XcovB[,-c(1)]

#add the propensity scores  
Xcov$pscore <- pscore3

#run genetic matching
m3_B <- GenMatch(dataB$treat, Xcov, BalanceMatrix=Xcov, 
               estimand="ATT", pop.size = 100, print=FALSE)

#create weights based on genetic matching
weightsm3_B <- get_gen_weights(dataB$treat, m3_B$matches)

#estimate treatment effect using the weights
est_3_B <- lm(Y ~ . + I(G1^2) + age:sex + famrel:traveltime + I(failures^2) 
              + I(health^2) + famrel:health, weights=weightsm3_B, data = dataB)
est_3_B <- summary(est_3_B)$coef["treat",1]
est_3_B

balance_func(XcovB, covsB, weightsm3_B)
```
   


#  Model fitting for World B: Ignoring the complex response surface   

Calculate p-scores   
```{r, cache=TRUE}
dat <- dataGen_B(data)
dataB <- dat[[1]]
outcomes <- dat[[2]]

pscore_fit4 <- glm(treat ~ ., data = dataB[,-c(1)], family = "binomial")
pscore4 <- predict(pscore_fit4, dataB, type = "response")

XcovB_sub <- XcovB[,-c(13:18)]
covsB_sub <- c("G1", "sex", "age", "romantic", "absences", "failures", "Walc", "traveltime", "famrel", "goout", "health")
```
   
IPTW weights   
```{r, cache=TRUE}
weightsm4_B <- get_IPTW_weights(dataB$treat, pscore4)
est_4_B <- lm(Y ~ ., weights=weightsm4_B, data = dataB)
est_4_B <- summary(est_4_B)$coef["treat",1]
est_4_B

balance_func(XcovB_sub, covsB_sub, weightsm4_B)
balance_func(XcovB, covsB, weightsm4_B)
```
    
Optimal matching  
```{r, cache=TRUE}
m5_B <- pairmatch(pscore_fit4)
weightsm5_B <- get_optm_weights(dataB$treat, m5_B)
est_5_B <- lm(Y ~ ., weights=weightsm5_B, data = dataB)
est_5_B <- summary(est_5_B)$coef["treat",1]
est_5_B

balance_func(XcovB_sub, covsB_sub, weightsm5_B)
balance_func(XcovB, covsB, weightsm5_B)
```
  
Genetic matching   
```{r, warning=FALSE, message=FALSE, cache=TRUE}
Xcov <- XcovB_sub[,-c(1)]

#re-save p score
Xcov$pscore <- pscore4

#run genetic matching
m6_B <- GenMatch(dataB$treat, Xcov, BalanceMatrix=Xcov, 
                 estimand="ATT", pop.size = 100, print=FALSE)

#create weights based on genetic matching
weightsm6_B <- get_gen_weights(dataB$treat, m6_B$matches)

#estimate treatment effect using the weights
est_6_B <- lm(Y ~ ., weights=weightsm6_B, data = dataB)
est_6_B <- summary(est_6_B)$coef["treat",1]
est_6_B

balance_func(XcovB_sub, covsB_sub, weightsm6_B)
balance_func(XcovB, covsB, weightsm6_B)
```    
     
Estimate PATT
```{r, cache=TRUE}
#calculate PATT
SATTs <- rep(NA, 100000)
for(i in 1:100000){
  dat <- dataGen_B(data)
  dataB <- dat[[1]]
  outcomes <- dat[[2]]
  SATTs[i] <- mean(outcomes$Y1[dataB$treat==1]) - mean(outcomes$Y0[dataB$treat==1])
}
PATT <- mean(SATTs)
PATT
```  
    
Verify that the treatment group is never larger than controls  
```{r}
ntreat <- rep(NA, 1000)
for(i in 1:1000){
  dataB <- dataGen_B(data)[[1]]
  ntreat[i] <- sum(dataB$treat==1)
}
sum((ntreat/nrow(dataB))>.5)
mean(ntreat/nrow(dataB))
```

Randomization distributions:    
```{r, cache=TRUE, message=FALSE, warning=FALSE}
nSim  <- 1000
simB_IPTW <- simB_IPTW_WrongMod <- rep(NA, nSim)
simB_optm <- simB_optm_WrongMod <- rep(NA, nSim)
simB_Gen <- simB_Gen_WrongMod <- rep(NA, nSim)


mat <- matrix(NA, ncol=length(covsB), nrow=nSim)
colnames(mat) <- covsB
simB_IPTW_bal <- simB_IPTW_WrongMod_bal <- mat
simB_optm_bal <- simB_optm_WrongMod_bal <- mat
simB_Gen_bal <- simB_Gen_WrongMod_bal <- mat

for(i in 1:1000){
  dataB <- dataGen_B(data)[[1]]
  
  pscore_fit3 <- glm(treat ~ . + I(G1^2) + age:sex + I(failures^2) 
                   + I(health^2) + famrel:traveltime + famrel:health, 
                   data = dataB[,-c(1)], family = "binomial")
  pscore3 <- predict(pscore_fit3, dataB[-1], type = "response")
  pscore_fit4 <- glm(treat ~ ., data = dataB[,-c(1)], family = "binomial")
  pscore4 <- predict(pscore_fit4, dataB[-1], type = "response")
  
  weightsm1_B <- get_IPTW_weights(dataB$treat, pscore3)
  est_1_B <- lm(Y ~ . + I(G1^2) + age:sex + I(failures^2) 
                   + I(health^2) + famrel:traveltime + famrel:health, 
                weights=weightsm1_B, data = dataB)
  simB_IPTW[i] <- summary(est_1_B)$coef["treat",1]
  simB_IPTW_bal[i,] <- balance_func(XcovB, covsB, weightsm1_B)$diff.m
  
  m2_B <- pairmatch(pscore_fit3)
  weightsm2_B <- get_optm_weights(dataB$treat, m2_B)
  est_2_B <- lm(Y ~ . + I(G1^2) + age:sex + I(failures^2) 
                   + I(health^2) + famrel:traveltime + famrel:health, 
                weights=weightsm2_B, data = dataB)
  simB_optm[i] <- summary(est_2_B)$coef["treat",1]
  simB_optm_bal[i,] <- balance_func(XcovB, covsB, weightsm2_B)$diff.m
  
  #save x covariates
  Xcov <- XcovB[,-c(1)]
  #add the propensity scores  
  #Xcov$pscore <- pscore3

  #run genetic matching
  m3_B <- GenMatch(dataB$treat, Xcov, BalanceMatrix=Xcov, 
               estimand="ATT", pop.size = 100, print=FALSE)
  weightsm3_B <- get_gen_weights(dataB$treat, m3_B$matches)
  est_3_B <- lm(Y ~ . + I(G1^2) + age:sex + I(failures^2) 
                + I(health^2) + famrel:traveltime + famrel:health, 
                weights=weightsm3_B, data = dataB)
  simB_Gen[i] <- summary(est_3_B)$coef["treat",1]
  simB_Gen_bal[i,] <- balance_func(XcovB, covsB, weightsm3_B)$diff.m
  
  #iptw weights
  weightsm4_B <- get_IPTW_weights(dataB$treat, pscore4)
  est_4_B <- lm(Y ~ ., weights=weightsm4_B, data = dataB)
  simB_IPTW_WrongMod[i] <- summary(est_4_B)$coef["treat",1]
  simB_IPTW_WrongMod_bal[i,] <- balance_func(XcovB, covsB, weightsm4_B)$diff.m
  
  #optimal matching
  m5_B <- pairmatch(pscore_fit4)
  weightsm5_B <- get_optm_weights(dataB$treat, m5_B)
  est_5_B <- lm(Y ~ ., weights=weightsm5_B, data = dataB)
  simB_optm_WrongMod[i] <- summary(est_5_B)$coef["treat",1]
  simB_optm_WrongMod_bal[i,] <- balance_func(XcovB, covsB, weightsm5_B)$diff.m
  
  #save selected covariates
  Xcov <- XcovB_sub[,-c(1)]
  #re-save p score
  #Xcov$pscore <- pscore4

  #run genetic matching
  m6_B <- GenMatch(dataB$treat, Xcov, BalanceMatrix=Xcov, 
                 estimand="ATT", pop.size = 100, print=FALSE)
  weightsm6_B <- get_gen_weights(dataB$treat, m6_B$matches)
  est_6_B <- lm(Y ~ ., weights=weightsm6_B, data = dataB)
  simB_Gen_WrongMod[i] <- summary(est_6_B)$coef["treat",1]
  simB_Gen_WrongMod_bal[i,] <- balance_func(XcovB, covsB, weightsm6_B)$diff.m
  
  print(i)
}

```    
    
Save everything, just in case!
```{r}
# statsB_final <-cbind(simB_IPTW,  simB_optm,  simB_Gen, simB_IPTW_WrongMod,
# simB_optm_WrongMod,simB_Gen_WrongMod)
# 
# colnames(statsB_final) <- c("simB_IPTW","simB_optm","simB_Gen",                   "simB_IPTW_WrongMod","simB_optm_WrongMod","simB_Gen_WrongMod")
# 
# write.csv(statsB_final, "./statsB_final.csv")
# write.csv(simB_IPTW_bal, "./IPTW_bal_B.csv")
# write.csv(simB_optm_bal, "./optm_bal_B.csv")
# write.csv(simB_Gen_bal, "./Gen_bal_B.csv")
# write.csv(simB_IPTW_WrongMod_bal, "./IPTW_WM_bal_B.csv")
# write.csv(simB_optm_WrongMod_bal, "./optm_WM_bal_B.csv")
# write.csv(simB_Gen_WrongMod_bal, "./Gen_WM_bal_B.csv")
```
    
Summary table  
```{r, cache=TRUE}
Biases <- round(abs(c(mean(simB_IPTW),
                     mean(simB_optm),
                     mean(simB_Gen),
                     mean(simB_IPTW_WrongMod),
                     mean(simB_optm_WrongMod),
                     mean(simB_Gen_WrongMod))-PATT),2)
RMSEs <- round(c(sqrt(mean((simB_IPTW-PATT)^2)),
                 sqrt(mean((simB_optm-PATT)^2)),
                 sqrt(mean((simB_Gen-PATT)^2)),
                 sqrt(mean((simB_IPTW_WrongMod-PATT)^2)),
                 sqrt(mean((simB_optm_WrongMod-PATT)^2)),
                 sqrt(mean((simB_Gen_WrongMod-PATT)^2))),2)
Included_Diff <- round(c(mean(abs(simB_IPTW_bal[,c(1:17)])),
                     mean(abs(simB_optm_bal[,c(1:17)])),
                     mean(abs(simB_Gen_bal[,c(1:17)])),
                     mean(abs(simB_IPTW_WrongMod_bal[,c(1:11)])),
                     mean(abs(simB_optm_WrongMod_bal[,c(1:11)])),
                     mean(abs(simB_Gen_WrongMod_bal[,c(1:11)]))),2)
Excluded_Diff <- round(c(mean(abs(simB_IPTW_bal[,c(1:17)])),
                     mean(abs(simB_optm_bal[,c(1:17)])),
                     mean(abs(simB_Gen_bal[,c(1:17)])),
                     mean(abs(simB_IPTW_WrongMod_bal[,c(12:17)])),
                     mean(abs(simB_optm_WrongMod_bal[,c(12:17)])),
                     mean(abs(simB_Gen_WrongMod_bal[,c(12:17)]))),2)

results <- data.frame(Bias = Biases,
                      RMSE = RMSEs,
                      Mean_Sd_Diff_Included = Included_Diff,
                      Mean_Sd_Diff_NotIncluded = Excluded_Diff)
rownames(results) <- c("IPTW - Correct Model", "Optimal - Correct Model", 
                       "Genetic - Correct Model",
                       "IPTW - Wrong Model", "Optimal - Wrong Model",
                       "Genetic - Wrong Model")
kable(results)
```    
    
Sampling distributions  
```{r, cache=TRUE}
par(mfrow = c(2,3))
hist(simB_IPTW, xlim=c(1.5,3.5), 
     main="IPTW Correct Model", xlab="ATT Estimate")
abline(v=PATT, lwd=2, col=4)
abline(v=mean(simB_IPTW), lwd=2, col=2, lty=2)
hist(simB_optm, xlim=c(1.5,3.5), 
     main="Opt Matching Correct Model", xlab="ATT Estimate")
abline(v=PATT, lwd=2, col=4)
abline(v=mean(simB_optm), lwd=2, col=2, lty=2)
hist(simB_Gen, xlim=c(1.5,3.5),
     main="Genetic Matching Correct Model", xlab="ATT Estimate")
abline(v=PATT, lwd=2, col=4)
abline(v=mean(simB_Gen), lwd=2, col=2, lty=2)
hist(simB_IPTW_WrongMod, xlim=c(1.5,3.5), 
     main="IPTW Wrong Model", xlab="ATT Estimate")
abline(v=PATT, lwd=2, col=4)
abline(v=mean(simB_IPTW_WrongMod), lwd=2, col=2, lty=2)
hist(simB_optm_WrongMod, xlim=c(1.5,3.5), 
     main="Opt Matching Wrong Model", xlab="ATT Estimate")
abline(v=PATT, lwd=2, col=4)
abline(v=mean(simB_optm_WrongMod), lwd=2, col=2, lty=2)
hist(simB_Gen_WrongMod, xlim=c(1.5,3.5),
     main="Genetic Matching Wrong Model", xlab="ATT Estimate")
abline(v=PATT, lwd=2, col=4)
abline(v=mean(simB_Gen_WrongMod), lwd=2, col=2, lty=2)
```  
   
Comparing to matching with replacement:  
```{r}
simB_wreplace
for(i in 1:nSim){
  dataB <- dataGen_B(data)[[1]]
  
  pscore_fit3 <- glm(treat ~ . + I(G1^2) + age:sex + I(failures^2) 
                   + I(health^2) + famrel:traveltime , 
                   data = dataB[,-c(1)], family = "binomial")
  pscore3 <- predict(pscore_fit3, dataB[-1], type = "response")
  
  match1 <- arm::matching(dataB$treat, score=pscore3, replace = TRUE)
  weightsm2_B <- rep(0, nrow(dataB))
  # weightsm2_B[which(dataB$treat==0) %in% match1$matched] <- 1
  weightsm2_B[dataB$treat==0] <- match1$cnts
  weightsm2_B[dataB$treat==1] <- 1
  
  est_2_B <- lm(Y ~ . + I(G1^2) + age:sex + famrel:traveltime + 
              famrel:health, weights=weightsm2_B, data = dataB)
  simB_wreplace[i] <- summary(est_2_B)$coef["treat",1]

}


mean(simB_wreplace)
```

  
Citation:  
P. Cortez and A. Silva. Using Data Mining to Predict Secondary School Student Performance. In A. Brito and J. Teixeira Eds., Proceedings of 5th FUture BUsiness TEChnology Conference (FUBUTEC 2008) pp. 5-12, Porto, Portugal, April, 2008, EUROSIS, ISBN 978-9077381-39-7. [Web Link]  

  
